---
title: "spotInRoi"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{spotInRoi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# What is spotInRoi analysis

spotInRoi (SIR) analysis is used to for:

1. Counting 3D spots that position within given 2D region of interests (ROIs).
2. Fetch 3D coordinate of spots and perform nearest neighbor distance (NND) calculation.

A three-step workflow is designed for this analysis:

1. Generate the spots and sample tables with [RNA-FISH pipeline][fish-analysis]. 
2. Draw 2D ROIs to generate the SIR table with the [Puncta Tracker imageJ plugin][puncta-tracker].
3. Count spots and (if needed) perform NND calculation with this package.

[fish-analysis]: https://github.com/yeyuan98/rna_fish_analysis
[puncta-tracker]: https://github.com/yeyuan98/punctaTracker

This vignette describes Step#3 above in detail. A small example dataset is 
included in this package.

# The example dataset

A sampled dataset is provided:

<pre>
├── fish-analysis
│   ├── exon.csv
│   ├── intron.csv
│   └── samples.xlsx
└── puncta-tracker
    ├── metadata.csv
    ├── spotInRoi_exonZ1X5I0.csv
    └── spotInRoi_intronZ3X5I0.csv
</pre>

Spots and sample tables are under the `fish-analysis` folder while 
SIR tables are under the `puncta-tracker` folder.

In the PunctaTracker analysis, we have drawn two ROIs for each cell. The first 
ROI of a cell always marks the nucleus boundary while the second always marks 
the cell.

# Workflow

Below I showcase the following tasks:

1. Count overall number of spots in each cell.
2. Count number of spots in the nucleus and whole cell ROIs.
3. Calculate NND to determine which exon spots colocalize with any intron.

First, define all relevant paths and load packages.

```{r}
path.root <- system.file("extdata/rna-fish-example", package = "ijAnalysis")
#   Paths of fish-analysis
path.fish <- file.path(path.root, "fish-analysis")
path.exon <- file.path(path.fish, "exon.csv")
path.intron <- file.path(path.fish, "intron.csv")
path.samples <- file.path(path.fish, "samples.xlsx")
#   Paths of puncta tracker
path.pt <- file.path(path.root, "puncta-tracker")
path.meta <- file.path(path.pt, "metadata.csv")
path.sirExon <- file.path(path.pt, "spotInRoi_exonZ1X5I0.csv")
path.sirIntron <- file.path(path.pt, "spotInRoi_intronZ3X5I0.csv")
```

```{r}
# Load package of course
library(ijAnalysis)
library(dplyr)
library(ggplot2)
```

Next, showcase Task 1 & 2, which are somewhat simple.

```{r}
# Task 1 - Count number of spots in each cell and plot
#   Load data and count
samples <- rfish_read_samples(path.samples)
dots.exon <- rfish_read_dots(path.exon)
dots.intron <- rfish_read_dots(path.intron)
df <- rbind(
  rfish_count(samples, dots.exon) |> mutate(probe = "exon"),
  rfish_count(samples, dots.intron) |> mutate(probe = "intron")
)
df <- df[df$num.cells > 0,]
#   Arrange order and plot
df$probe <- ordered(df$probe, levels = c("exon", "intron"))
#     dataset have two samples CT4 and CT16
df$sample <- ordered(df$sample, levels = c("CT4", "CT16"))
#     plot X = sample, Y = dots-per-cell (dpc), color fill group = probe
ctcf_plot(df, sample, dpc, probe)+
  scale_y_continuous(limits = c(0,30), expand = c(0,.01))+
  xlab("Time Point")+
  ylab("Spots/cell")
```

```{r}
# Task 2 -  Count number of exon spots in ROIs.
#   read SIR table
exon <- spotInRoi_read_csv(path.sirExon)
#   count by cell
exon <- exon |> 
  group_by(tp, meas, plane) |> 
  # get number of spots and percent nuclear
  summarize(
    n.nuc = spot.count[1], n.tot = spot.count[2], 
    perc.nuc = n.nuc/n.tot*100, .groups = "drop")
#   filter (must have valid data and at least 3 dots in cell ROI)
exon <- exon[complete.cases(exon) & exon$n.tot >= 3,]
#   plot X = tp, Y = perc.nuc (percent nuclear)
exon$tp <- ordered(exon$tp, levels = c("CT4", "CT16"))
ctcf_plot(exon, tp, perc.nuc)+
  scale_y_continuous(limits = c(0, 105), expand = c(0,.01))+
  xlab("Time Point")+
  ylab("Perc. of nuclear exon spots/cell")
```

Task 3 is more complicated. We start by calculating NND values in pixel unit:

```{r}
# Task 3 - Calculate NND for each exon spot to intron spots.
#   read SIR tables
exon <- spotInRoi_read_csv(path.sirExon)
intron <- spotInRoi_read_csv(path.sirIntron)
#   select spots (here, select spots in the nucleus)
exon <- spotInRoi_selectNCT(exon, nucleus)
intron <- spotInRoi_selectNCT(intron, nucleus)
#   remove entries that show no exon spots
exon <- exon |> filter(spot.count > 0)
#   join two SIR tables
exon2intron <- inner_join(
  x = exon |> select(tp, meas, plane, spotIn),
  y = intron |> select(tp, meas, plane, spotIn),
  by = c("tp", "meas", "plane")
)
#   compute NND
nnd_compute <- 
  # for each exon spot, take intron spots in 3 Z-frame proximity.
  \(x,y) ij_nnd.compute(x, y, z.proximal = 3)
exon2intron$result <- 
  with(exon2intron, purrr::map2(spotIn.x, spotIn.y, nnd_compute))
#   (optionally) sort spots by NND value in each cell (increasing order)
exon2intron <- spotInRoi_nnd.sort(exon2intron)
#   expand NND result table
#     before: each row is a cell
#     after: each row is a exon spot (w/ its NND values)
exon2intron <- spotInRoi_as.data.frame.df.nnd(exon2intron)
```

Now, `exon2intron` is a data frame where each row is a exon spot with NND 
intron values. However, the NND values `$nnd` are in pixel unit.

To convert to physical unit, we need to fetch the physicalSizes columns that 
are present only in the RNA-FISH pipeline outputs 
(tables under folder `fish-analysis`):

```{r}
# Fetch physicalSizes columns
exon2intron <- spotInRoi_addMeta(
  df = exon2intron,
  idMap = read.csv(path.meta),
  meta = rfish_read_samples(path.samples),
  starts_with("phys") # select physicalSizes columns
)
exon2intron$nnd <- exon2intron$nnd * exon2intron$physX * 1E3 # unit = nm
exon2intron$colocalized <- exon2intron$nnd < 400 # cutoff = 400nm
# Compute percent of exon spots colocalized to any intron
exon2intron <- exon2intron |> 
    group_by(tp, meas, plane) |> 
    summarize(
      n.total = n(), n.coloc = sum(colocalized), 
      perc.coloc = n.coloc / n.total*100, .groups = "drop")
# Plot, X = tp, Y = perc.coloc
ctcf_plot(exon2intron, tp, perc.coloc)+
  scale_y_continuous(limits = c(0,105), expand = c(0,.01))+
  xlab("Time Point")+
  ylab("Percent of colocalized exon spots/cell")
```

Notes:

1. `z.proximal` parameter of `ij_nnd.compute` controls how NND is identified. 
For a real dataset that adopted the `Z1X5I0` for exon and `Z3X5I0` for intron, 
usually we use `z.proximal = 1.5` to find introns within 1.5 Z-frames of the 
exon for calculating NND values in the XY plane. Here, `z.proximal = 3` is used 
because the dataset is sampled and for some exon spots no intron spot could be 
found within 1.5 Z-plane proximity.
